---
title: "Mini project 1 - Marathon Results  "
author: "Prathyusha Bhuma - pbhuma7210@floridapoly.edu"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

```{r}

#Loading libraries
library(tidyverse)
library(ggplot2)
```


```{r}
# Loading data
marathon_results<-read_csv("/Users/prathyushabhuma/Documents/Florida Polytechnic University/Data Visualization and Reproducible Research/dataviz_final_project/data/marathon_results_2017.csv")
```
```{r}
#dimensions 
dim(marathon_results)

```

```{r}
head(marathon_results)
```
```{r}
# Count the number of Finishers by Gender
Finishers_by_gender <- table(marathon_results$`M/F`)
```


```{r}
Finishers_by_gender
```



```{r}
# Convert the Finishers count to a data frame
Finishers_df <- as.data.frame(Finishers_by_gender)

# Rename the columns for clarity
colnames(Finishers_df) <- c("Gender", "Count")
```

```{r}
Finishers_df
```

```{r}
# Create a pie chart
ggplot(Finishers_df, aes(x = "", y = Count, fill = Gender)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(title = "Distribution of Finishers by Gender") +
  scale_fill_manual(values = c("blue", "pink")) +  # Customizing fill colors for genders
  theme_void()  # Removing unnecessary elements like axes and grid lines

```
```{r}
# Group the data by country and calculate the average finishing time for each country
country_avg_time <- marathon_results %>%
  group_by(Country) %>%
  summarise(avg_time=mean(`Official Time`))

# Sort the countries based on the average finishing time in Ascending order
sorted_countries <- country_avg_time %>%
  arrange(avg_time)
sorted_countries
```


```{r}
# Select the top 5 countries with the fastest runners
top_5_countries <- head(sorted_countries, 5)

#Create a bar plot for the top 5 countries with the fastest runners
ggplot(top_5_countries, aes(x = reorder(Country, avg_time) , y = avg_time )) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Top 5 Countries with the Fastest Runners",
       x = "Country",
       y = "Average Finishing Time") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme_minimal()
```

```{r}

# Count the number of Finishers from each country
Finishers_by_country <- marathon_results %>%
  group_by(Country) %>%
  summarise(Num_Finishers = n()) 


# View the count of participants from each country
print(Finishers_by_country)
```

```{r}
# Select the top 10 countries with the More Number of Participants
top_10_countries<-arrange(Finishers_by_country, desc(Num_Finishers))
top_10_countries
```

```{r}
Top_10<-head(top_10_countries,10)
Top_10
```


```{r}
# Create a lollipop chart
ggplot(Top_10,
       aes(x = Country, y = Num_Finishers , 
           color = Country)) +
  geom_pointrange(aes(ymin = 0, ymax = Num_Finishers)) +
  guides(color = FALSE) +
  labs(title = "Number of Finishers by Country",
       x = "Country",
       y = "Number of Finishers" )+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

```
```{r}
# Create a histogram to visualize the distribution of Finishers ages
ggplot(marathon_results, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Finishers Ages",
       x = "Age",
       y = "Number of Finishers") +
  theme_minimal()

```



```{r}
# Create a boxplot to visualize the distribution of Finishers ages by gender
ggplot(marathon_results, aes(x =`M/F`,y = Age, fill =`M/F`)) +
  geom_boxplot(alpha = 0.3) +
  labs(title = "Age Distribution based on Gender",
       x = "Gender",
       y = "Age") +
  theme_minimal()
```

```{r}

# Extension to the original mini project that was submitted earlier with a new visualization
ggplot(marathon_results, aes(x = Age, y = `Official Time`, color = `M/F`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  labs(title = "Age vs. Official Finishing Time",
       x = "Age",
       y = "Official Finishing Time (seconds)",
       color = "Gender") +
  scale_color_manual(values = c("M" = "blue", "F" = "pink")) +
  theme_minimal()
```

Comment: The scatter plot enables us to explore the relationship between the official finishing times and the ages of the marathon participants. This can provide insights into whether certain age groups tend to finish faster or slower on average. Also, it will allow us to visualize how the finishing times vary with age and whether there are noticeable differences between genders in this aspect.